Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-12-07 article.tex
asd

\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{chains}
\usetikzlibrary{angles, quotes}
\usetikzlibrary{decorations.pathreplacing}
\usepackage{graphicx} % Required for inserting images
\usepackage[T1]{fontenc}
\usepackage{beramono}
\usepackage{courier}
\usepackage{listings}
\usepackage{xpatch}
\usepackage{realboxes}
\usepackage{titlesec}
\usepackage{xcolor}

@@

@@

Compilation refers to the process of translating from one language to another, most often from a high-level programming language intended for humans to work with, to machine- or bytecode intended to be executed on a target platform. This process can be divided into several distinct phases, which are grouped into one of two stages colloquially referred to as the \textit{frontend} and \textit{backend}. The former translating a high-level programming language to an \textit{intermediate representation} (IR) layer and the latter translating IR to executable machine code on a target platform or bytecode for a target \textit{virtual machine} (VM).

\begin{figure}[h]
  \centering
  \resizebox{0.8\textwidth}{!}{
\begin{tikzpicture}
  \tikzset{
box/.style={draw,
  inner sep=0.3cm,
    minimum height=1.75cm,
    align=center},
}

@@

  \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=4ex}]
    (0.3,-1) -- (0.3,1) node[midway,xshift=3em,rotate=-90]{Backend};
  \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=4ex}]
    (0.3,1.25) -- (0.3,3.25) node[midway,xshift=3em,rotate=-90]{Frontend};
\end{tikzpicture}}
\caption{Compiler phases, backend highlighted}
\end{figure}
Most operations of a general-purpose programming language are translated to a set of logic,  control, and arithmetic instructions to be executed sequentially on a computer processor: a single circuit/chip, referred to as the \textit{central processing unit} (CPU), the design of which has varied and evolved over time. Most processors executing logic operations are \textit{register machines}, in that they use a limited set of \textit{general-purpose registers} (GPRs) to store working values in combination with \textit{random access memory} (RAM) for mid-term, and other I/O peripherals for long-term storage.

This can largely be attributed to performance, as register machines routinely outperform so-called \textit{stack machines} \parencite{ShiYunhe2008VmsS} that are often used in VMs.
Because of the limited amount of GPRs available simultaneously, a crucial part of the backend stage for an optimizing compiler is assigning each variable  of the source program to a GPR in such a way that maximizes performance without sacrificing correctness.

@@

\section{Review of Literature}

The two primary sources are the \textit{Modern Compiler Implementation in ML} \cite{tiger} and Dragonbook \cite{dragon}, in addition to scientific publications concerning the linear scan approach.

\subsection{Control Flow Analysis}

The backend of a compiler takes some form of IR as input, usually a linear sequence of instructions for each separate function. This representation is close to the level of an actual processor by design, but it isn't immediately useful for the further analysis steps needed to generate optimized code for the target architecture. %%(FIXME: why not?)
The control flow of a given program refers to the order in which instructions are executed. While the flow of most instructions is linear, in the sense that the next instruction to be executed is located immediately after, some transfer the flow of execution elsewhere or terminate it. %% FIXME: why is this important?

@@

Such contiguous sequences of instructions are referred to as \textit{basic blocks}: a basic block is defined as a sequence of instructions with no branches in or out except for the first and last instructions, and can be collectively construed of as a node in a control flow graph, whose directed entry edges are from any block that directs control flow towards itself.

These instructions are referred to as \textit{terminators}, as they potentially terminate a contiguous flow of execution. This subset of instructions can be further divided into \textit{branch instructions} and those concerned with handling the callstack in function/subroutine management.

@@

hence a \textit{control flow graph} is derived from these. A terminator will branch elsewhere to continue execution. The destination of this has to be an annotated by a label, hence labels \textit{initiate} blocks.

\subsection{Liveness Analysis}

\subsection{Graph Coloring}

\subsection{Linear Scan}

@@

@@

%* list of methods for coalescing/allocating:
%   * ocamlgraph builtin
%   * greedy 
%   * briggs
%   * george
%   * welsh-powell?
\section{Graph Coloring}
The heuristic for graph coloring as introduced in both the Appel and Aho et al. texts is a linear time approximation of an NP-complete problem. It is based on an iterative approach.

@@

@@

\subsection{Benchmarking}

@@

>>>>

\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{angles}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{automata}
\usetikzlibrary{calc}
\usetikzlibrary{chains}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{positioning}
\usetikzlibrary{quotes}
\usepackage{graphicx} % Required for inserting images
\usepackage[T1]{fontenc}
\usepackage{beramono}
\usepackage{courier}
\usepackage{listings}
\usepackage{xpatch}
\usepackage{realboxes}
\usepackage{titlesec}
\usepackage{xcolor}

@@

\tikzset{
box/.style={draw,
  inner sep=0.3cm,
    minimum height=1.75cm,
    align=center},
cbox/.style={draw,
  font=\ttfamily,},
}
\tikzset{invisible/.style={minimum width=0mm,inner sep=0mm,outer sep=0mm}}

@@

Compilation refers to the process of translating from one language to another, most often from a high-level programming language intended for humans to work with, to machine- or bytecode intended to be executed on a target platform. This process can be divided into several distinct phases, which are grouped into one of two stages colloquially referred to as the \textit{frontend} and \textit{backend}. The former translating a high-level programming language to an \textit{intermediate representation} (IR) and the latter translating IR to executable machine code of a target architecture or bytecode of a target \textit{virtual machine} (VM).

\begin{figure}[h]
  \centering
  \resizebox{0.8\textwidth}{!}{
\begin{tikzpicture}

@@

  \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=4ex}]
    (0.3,-1) -- (0.3,1) node[midway,xshift=3em,rotate=-90]{Backend};
  \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=4ex}]
    (0.3,1.25) -- (0.3,3.25) node[midway,xshift=3em,rotate=-90]{Frontend};
\end{tikzpicture}}
\caption{Compiler phases, backend highlighted}
\end{figure}
Most operations of a general-purpose programming language are translated to a set of control, logic, and arithmetic instructions to be executed sequentially on a computer processor: a single circuit/chip, referred to as the \textit{central processing unit} (CPU), the design of which has varied and evolved over time.

Most CPUs are \textit{register machines}, in that they use a limited set of \textit{general-purpose registers} (GPRs) to store working values in combination with \textit{random access memory} (RAM) for mid-term, and other I/O peripherals for long-term storage. This can largely be attributed to performance, as register machines routinely outperform \textit{stack machines} \parencite{ShiYunhe2008VmsS} that are often used in VMs.
Because of the limited amount of GPRs available simultaneously, a crucial part of the backend stage for an optimizing compiler is assigning each variable  of the source program to a GPR in such a way that maximizes performance without sacrificing correctness.

@@

%\section{Review of Literature}

%The two primary sources are the \textit{Modern Compiler Implementation in ML} \cite{tiger} and Dragonbook \cite{dragon}, in addition to scientific publications concerning the linear scan approach.

% TODO: figure out what traces are

The backend of a compiler takes some form of IR as input, usually a linear sequence of instructions for each separate function. This representation is close to the level of an actual processor by design, but it isn't immediately useful for the further analysis steps needed to generate optimized code for the target architecture. %%(FIXME: why not?)
The control flow of a given program refers to the order in which instructions are executed. While the flow of most instructions is linear, in the sense that the next instruction executed is located immediately after, some transfer the flow of execution elsewhere or even terminate it. %% FIXME: why is this important?

@@

Continuous sequences of instructions are referred to as \textit{basic blocks}. A basic block is defined as a sequence of instructions with no branches in or out except for the first instruction, which follows immediately either the function entry point or a label,  and the last instruction, which either terminates or transfers the flow of execution, hence these are often referred to as terminators and can be thought of as a different categoriy of instructions entirely.

Basic blocks represent a single node in a \textit{control flow graph} (CFG), which is a directed graph whose edges denote transfer of  control flow. The unconditional branch terminator always transfers control flow to the block labelled, meaning only one successor will follow, whereas conditional branching could transfer to either of the two, but because control flow analysis is not concerned with data, it simply adds both as successors. % TODO: rets 
While each block has 0-2 immediate successors, the amount of predecessors is unbounded.

\begin{figure}[h]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \begin{tikzpicture}
    \node [cbox, font=\ttfamily\tiny, align=left] (cfa) at (0,0) {define i32 @square(i32 \%0) \{\\1:\\\ \%2 = icmp sgt i32 \%0, 0\\\ br i1 \%2, label \%3, label \%9\ ; succs = \%3, \%9\\3:\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ ; preds = \%1, \%3\\\ \%4 = phi i32 [ \%0, \%1 ], [ \%7, \%3 ]\\\ \%5 = phi i32 [ 0, \%1 ], [ \%6, \%3 ]\\\ \%6 = add i32 \%5, \%0\\\ \%7 = sub i32 \%4, 1\\\ \%8 = icmp ugt i32 \%7, 0\\\ br i1 \%8, label \%3, label \%9\ ; succs = \%3, \%9\\9:\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ ; preds = \%3, \%1\\\ \%10 = phi i32 [ 0, \%1 ], [ \%6, \%3 ]\\\ ret i32 \%10\\\}};
  \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \begin{tikzpicture}
    \node [cbox, font=\ttfamily\tiny, align=left] (a) at (0,0) {1:\\\ \%2 = icmp sgt i32 \%0, 0\\\ br i1 \%2, label \%3, label \%9\ \ \ \ \ \ \ };
    \node [cbox, font=\ttfamily\tiny, align=left, below=0.5cm of a] (b) {3:\\\ \%4 = phi i32 [ \%0, \%1 ], [ \%7, \%3 ]\\\ \%5 = phi i32 [ 0, \%1 ], [ \%6, \%3 ]\\\ \%6 = add i32 \%5, \%0\\\ \%7 = sub i32 \%4, 1\\\ \%8 = icmp ugt i32 \%7, 0\\\ br i1 \%8, label \%3, label \%9\\9:\\\ \%10 = phi i32 [ 0, \%1 ], [ \%6, \%3 ]\\\ ret i32 \%10\\\}};
    \node [cbox, font=\ttfamily\tiny, align=left, below=0.5cm of b] (c) {9:\\\ \%10 = phi i32 [ 0, \%1 ], [ \%6, \%3 ]\\\ ret i32 \%10};
    \node[right=2cm, invisible] (ab) at ($(a.south)!0.5!(b.north)$) {};
    \node[right=2cm, invisible] (bc) at ($(b.south)!0.5!(c.north)$) {};
  \draw [-latex] (a) -- (b);
  %\draw [-latex] (a) to[out=-15,in=15] (c);
  \draw [-latex] (b) -- (c);
  %\draw [-latex] (c.north) to[bend right=1000] (b.north);
  \draw  (a.south) to [in=0] (ab);
  \draw  (ab) to [in=0] (bc);
  \draw [-latex]  (bc) to [out=180,in=70] (c.north);
  \end{tikzpicture}
  \end{subfigure}
  \hfill
  \caption{Compiler phases, backend highlighted}
\end{figure}

%These instructions are referred to as \textit{terminators}, as they potentially terminate a contiguous flow of execution. This subset of instructions can be further divided into \textit{branch instructions} and those concerned with handling the callstack in function/subroutine management.

@@

%hence a \textit{control flow graph} is derived from these. A terminator will branch elsewhere to continue execution. The destination of this has to be an annotated by a label, hence labels \textit{initiate} blocks.

@@

% * Different approaches:
%   * Naive/greedy 
%   * Dataflow analysis 
%     * forward must
%     * backwards may
%   * linear scan

The purpose of liveness analysis is to determine which variables are live at which program point, which in turn is used to

@@

%* list of methods for coalescing/allocating:
%   * ocamlgraph builtin
%   * greedy 
%   * briggs
%   * george
%   * welsh-powell?
\section{Graph Coloring}
The heuristic for graph coloring as introduced in Appel and Aho et al. texts is a linear time approximation of an NP-complete problem. It is based on an iterative approach.

@@

As linear scan

@@

\newpage
\subsection{Benchmarking}

@@

%% NOTE: måske noget om hvordan intel laver sin egen arkitektur?
